Online Updating of Variational Bayes for Heterogenous Forecasts of Vehicle Trajectory

### 2018-03-06 Cars/paper.tex
section six editing

A crucial aspect of the application of the discussed time series models is through the relationship between the posterior distribution formed from observations relatting to the first $N$ vehicles and the parameter vector for the additional new vehicle, $\theta_{N+1}$. Under the homogenous model $\theta$ is shared across all vehicles and hence $\theta_{i} \equiv \theta$. In this case, if $N$ is large $p(\theta | \textbf{z}_{1:N}, z_{i, 1:T})$ will not vary much from $p(\theta | \textbf{z}_{1:N})$. Figure (\ref{fig:homogPosterior}) displays plots of the marginal posterior distributions associated with each element of $\theta$, with $p = q = 2$ resulting from $N = 2000$ vehicles. Arguably each of these marginal posterior distributions are very precise, and are unlikely to change much if updated using information from observations from the $(N+1)^{st}$ vehicle.

@@

Choosing the number of samples per estimate, $M$, involves a trade-off between the computation time per gradient estimate and the stochastic noise present in each estimate. An increased value of $M$ will the Monte-Carlo error, and generally reduce the number of iterations required for the ELBO to converge, at a linear increase in computation time per iteration. To reduce the Monte Carlo error, following \citet{Gunawan2017}, Randomised Quasi Monte Carlo (RQMC) is used instead, which has shown to be more efficient than standard Monte-Carlo in many applications (see \cite{{Niederreiter1992, Caflisch1998}). In RQMC a deterministic sequence of numbers in the unit hypercube (of dimensionality equal to $\theta_i$) are generated evenly, in the sense that they have a low star discrepency \citep{Sobol1976}, according to a Sobol Sequence \citep{Sobol1967}. These numbers are then randomised using the scrambled net method \citep{Matousek1998}, which preserves the low discrepency properties while inducing a central limit theorem. $M$ of the resulting numbers are transformed using the inverse-CDF of $q_{\lambda}(\theta | z)$ or $r(\epsilon)$ to simulate $M$ draws from these distributions.

@@

After obtaining the posterior distributions for each model conditioned on the training set of 2000 vehicles, with $p = q = 2$ and $K = 6$, forecasts are created and evaluated on the  further 1387 vehicles in the test set. For each of these vehicles $i > 2000$, after each of $T = 100, 110, \dots, 450$ observations, and at every forecast horizon $h = 1, \dots, 30$, point estimates of $z_{i, T+h}$ are obtained from every naive model. The time-series models are used to provide forecast densities of $p(z_{i, T+h} | \textbf{z}_{1:2000}, z_{i, 1:T})$, and point estimates using the coordinate pair $\{\hat{x}_{i, t}^*, \hat{y}_{i, t}^*\}$ which maximises this density. As data is observed every 100 milliseconds, this equates to forecasting the next three seconds of movement for every vehicle after every second of observation. The forecast density for the homogenous model is created using the MCMC samples of $p(\theta | \textbf{z}_{1:2000}, z_{i, T})$, while the IH and CH models produce $p(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ before creating forecasts through three methods:
\begin{enumerate}
\item Posterior sampling with RWMH-MCMC,
\item VB fitting $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ to the complete $z_{i, 1:T}$,
\item UVB fitting $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ using $q_{\lambda_S}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:S})$ and data $z_{i, S+1:T}$ as described in Section \ref{sec:UVB}.
\end{enumerate}

@@

Nine naive forecast models are provided where forecasts of $a_{i, t}$ and $\delta_{i, t}$ at each time $T$ are constants equal to $\hat{a}_{i, T}$ and $\hat{\delta}_{i, T}$ for $t = T+1, \dots, T+H$. The classification of each naive model is provided in Table \ref{tableNaive} as a combination of one estimator for $\hat{a}_{i, T}$ and one estimator for $\hat{\delta}_{i, T}$. Acceleration estimators are the most recent acceleration, $\hat{a}_{i, T} = a_{i, T}$, the average acceleration over the most recent second, $\hat{a}_{i, T} = 0.1 \sum_{s = T-9}^T a_{i, s}$, and zero acceleration $\hat{a}_{i, T} = 0$, equivalent to constant velocity. Similarly, steering angle estimators are the most recent angle, $\hat{\delta}_{i, T} = \delta_{i, T}$, the average angle over the most recent second, $\hat{\delta}_{i, T} = 0.1 \sum_{s = T - 9}^T \delta_{i, s}$, and $\hat{\delta}_{i, T} = \pi/2$, corresponding to the vehicle to driving directly forward.
\begin{table}
\begin{center}
\resizebox{\textwidth}{!}{%
\begin{tabular}{|l|c|c|c|}
\hline
& $\hat{\delta}_{i, T} = \delta_{i, T}$ & $\hat{\delta}_{i, T} = 1/10 \sum_{s=T-9}^T \delta_{i, s}$ & $\hat{\delta}_{i, T} = \pi/2$ \\
\hline
$\hat{a}_{i, T} = a_{i, T}$ & Naive 1 & Naive 2 & Naive 3 \\
$\hat{a}_{i, T} = 1/10\sum_{s=T-9}^T a_{i, s}$ & Naive 4 & Naive 5 & Naive 6\\
$\hat{a}_{i, T} = 0$ & Naive 7 & Naive 8 & Naive 9 \\
\hline
\end{tabular}}
\end{center}
\caption{Classification of naive forecast models by future acceleration and steering angle estimators.}
\label{tableNaive}
\end{table}

@@

Eucliedean Error for each model, averaged across test set vehicles at forecast horizons of one, two, and three seconds ($h = 10, 20$ and $30$) are provided in Figure \ref{fig:PredError}. Naive model performance is split into three groups according to the choice of acceleration forecast, and for a given $\hat{a}_{i, T}$ the choice of $\hat{\delta}_{i, T}$ has not substantially changed the forecast error. Zero acceleration naive models (Naive 7, 8, and 9) perform the best, followed by the averaged acceleration models (Naive 4, 5, and 6) and finally the current acceleration models (Naive 1, 2, and 3). At all forecast horizons the error for each implementation of each of the time series models are significantly lower than every naive model. The average forecast Euclidean Error for the best performing Naive model, Naive 8, and each of the time series models at each of these three forecast horizons are presented in Table \ref{table:meanError}.
\\

Figure (\ref{fig:PredErrorZ}) truncates Figure (\ref{fig:PredError}) to include only the time series models, each of which produces a forecast with a similar amount of error that does not increase substantially as the forecast horizon increases. This implies that there is no systematic increase or decrease in MAP forecast accuracy for choosing either the homogenous or introducing heterogeneity with the IH and CH models, however the UVB inference implementation of the IH model has increased the forecast error relative to MCMC and standard VB. The UVB implementation of the CH model produces forecast errors that are competitive with MCMC and standard VB, and so inference discussion will focus on the CH model rather than the IH model. 
\\

@@

The CH model assigns each vehicle in the test set one of a wide range of variances, while the homogenous model must use the same variance paramter for every vehicle. Changes in variance parameters has little impact on the location of the MAP of $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$, but impacts the level of uncertainty in the forecast density, causing systematic differences in the level of forecast uncertainty measured by logscores.
\\
Figure (\ref{fig:posVarMean}) displays plots of density of posterior means for each of acceleration and steering angle resulting from RWMH-MCMC sampling with $T = 450$. The homogenous model posterior means are indicated with vertical red lines. As in Figure \ref{fig:HierSingleKDE}, there is a large amount of variation in the distribution of variances for the test set, many of which are lower than the homogenous model: the homogenous posterior means are greater than $80\%$ and $91.5\%$ of CH posterior means for acceleration and the steering angle respectivley. 
\\

Figure \ref{fig:varSplit} shows a detailed breakdown of the median increase in predictive logscore obtained from using the CH model with UVB inference over the homogenous model at the three second forecast horizon. Vehicles are split into bivariate quintiles according to their posterior variance means for acceleration along the x-axis and angle along the y-axis. The CH model trajectory forecasts for drivers in the lower variance quintiles benefit the most from including heterogeneity. However the median difference is zero, or even negative, for forecasts of drivers in the fifth quintiles. This paper argues that heterogenous modelling is still useful in this scenario, as the individual vehicle variance estimates allow identification of vehicles with high variance, and potentially more erratic and dangerous driving styles.
\\

@@

The benefits of heterogenous modelling is only useful if the posterior distribution $p(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$, or its variational approximation $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$ can be inferred in the short time frames demanded by self-driving vehicles. Figure (\ref{fig:timing}) demonstrates the average computation time required for standard VB (Black) and UVB (coloured), where UVB is first fit to $100$ observations then repeatedly updated every $T-S$ observations for different values of $T-S$. The compuation time for standard VB increases linearly with $T$, while UVB is constant; posterior inference available at time $T$ utilising UVB can be conditioned on more data than standard VB. Parallelising the $M$ Monte-Carlo samples per iteration can further reduce computation time and make UVB feasible for more complex models. 
\\

As UVB involves replacing the posterior distribution at update with its variational approximation, the error introduced by UVB depends on how closely the variational approximation can represent the true posterior distribution. The left panel for Figure (\ref{fig:updateCost}) displays boxplots of the difference in logscore between UVB and standard VB at the three second forecast horizon for the CH model, and demonstrates that UVB in this case does not decrease the forecast logscores, as the difference between the logscores obtained by UVB and standard VB are distributed close to zero for several ranges of $T$. The right panel displays analogous boxplots for the IH model, where the UVB logscores tend to be lower than the standard VB logscores, matching the increased Euclidean error found for the IH model with UVB inference in Figure \ref{fig:PredErrorZ}. An explanation for this is given by the form of the approximating distribution used for the IH model, the multivariate normal distribution may not adequately approximate the posterior distribution, while the extra flexibilitiy of the mixture multivariate normal distribution for the CH model was sufficient to replace the posterior.

@@

\section{Discussion}
\label{sec:disc}
This paper demonstrates that the use of auto-regressive models time-series models drastically improves trajectory forecasts when compared to the competing models.  These models parameterise the dynamics of a driver's actions, with a range of parameter posterior means providing evidence of heterogeneity between the drivers of different vehicles. Incorpororating this heterogeneity through a hierarchical model has improved the forecast logscores, and hence reduced uncertainty, but requires an approximate inference scheme implemented with UVB. 
\\

>>>>

A crucial aspect of the application of the discussed time series models is through the relationship between the posterior distribution formed from observations relatting to the first $N$ vehicles and the parameter vector for the additional new vehicle, $\theta_{N+1}$. Under the homogenous model $\theta$ is shared across all vehicles and hence $\theta_{i} \equiv \theta$. In this case, if $N$ is large $p(\theta | \textbf{z}_{1:N}, z_{i, 1:T})$ will not vary much from $p(\theta | \textbf{z}_{1:N})$. Figure \ref{fig:homogPosterior} displays plots of the marginal posterior distributions associated with each element of $\theta$, with $p = q = 2$ resulting from $N = 2000$ vehicles. Arguably each of these marginal posterior distributions are very precise, and are unlikely to change much if updated using information from observations from the $(N+1)^{st}$ vehicle.

@@

Choosing the number of samples per estimate, $M$, involves a trade-off between the computation time per gradient estimate and the stochastic noise present in each estimate. An increased value of $M$ will the Monte-Carlo error, and generally reduce the number of iterations required for the ELBO to converge, at a linear increase in computation time per iteration. To reduce the Monte Carlo error, following \citet{Gunawan2017}, Randomised Quasi Monte Carlo (RQMC) is used instead, which has shown to be more efficient than standard Monte-Carlo in many applications (see \cite{Niederreiter1992, Caflisch1998}). In RQMC a deterministic sequence of numbers in the unit hypercube (of dimensionality equal to $\theta_i$) are generated evenly, in the sense that they have a low star discrepency, according to a Sobol Sequence \citep{Sobol1967}. These numbers are then randomised using the scrambled net method \citep{Matousek1998}, which preserves the low discrepency properties while inducing a central limit theorem. $M$ of the resulting numbers are transformed using the inverse-CDF of $q_{\lambda}(\theta | z)$ or $r(\epsilon)$ to simulate $M$ draws from these distributions.

@@

After obtaining RWMH-MCMC samples of the posterior distributions for each model conditioned on the training set of 2000 vehicles, with $p = q = 2$ and $K = 6$, forecasts are created and evaluated on the  further 1387 vehicles in the test set. For each of these vehicles $i > 2000$, after each of $T = 100, 110, \dots, 450$ observations, and at every forecast horizon $h = 1, \dots, 30$, point estimates of $z_{i, T+h}$ are obtained from every naive model. The time series models are used to provide forecast densities of $p(z_{i, T+h} | \textbf{z}_{1:2000}, z_{i, 1:T})$, and point estimates using the coordinate pair $\{\hat{x}_{i, t}^*, \hat{y}_{i, t}^*\}$ which maximises this density. As data is observed every 100 milliseconds, this equates to forecasting the next three seconds of movement for every vehicle after every second of observation. The forecast density for the homogenous model is created using the samples of $p(\theta | \textbf{z}_{1:2000}, z_{i, T})$, while the IH and CH models produce $p(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ before creating forecasts through three methods:
\begin{enumerate}
\item Posterior sampling with RWMH-MCMC,
\item VB fitting $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ to the complete $z_{i, 1:T}$,
\item UVB fitting $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ using $q_{\lambda_S}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:S})$ and data $z_{i, S+1:T}$ as described in Section \ref{sec:UVB}.
\end{enumerate}

@@

Nine naive forecast models are provided where forecasts of $a_{i, t}$ and $\delta_{i, t}$ at each time $T$ are constants equal to $\hat{a}_{i, T}$ and $\hat{\delta}_{i, T}$ for $t = T+1, \dots, T+H$. The classification of each naive model is provided in Table \ref{tableNaive} as a combination of one estimator for $\hat{a}_{i, T}$ and one estimator for $\hat{\delta}_{i, T}$. Acceleration estimators are the current acceleration, $\hat{a}_{i, T} = a_{i, T}$, the average acceleration over the previous second, $\hat{a}_{i, T} = 0.1 \sum_{s = T-9}^T a_{i, s}$, and zero acceleration $\hat{a}_{i, T} = 0$, equivalent to constant velocity. Similarly, steering angle estimators are the current angle, $\hat{\delta}_{i, T} = \delta_{i, T}$, the average angle over the previous second, $\hat{\delta}_{i, T} = 0.1 \sum_{s = T - 9}^T \delta_{i, s}$, and $\hat{\delta}_{i, T} = \pi/2$, corresponding to the vehicle to driving directly forward.
\begin{table}
\begin{center}
\resizebox{\textwidth}{!}{%
\begin{tabular}{|l|c|c|c|}
\hline
& $\hat{\delta}_{i, T} = \delta_{i, T}$ & $\hat{\delta}_{i, T} = 0.1 \sum_{s=T-9}^T \delta_{i, s}$ & $\hat{\delta}_{i, T} = \pi/2$ \\
\hline
$\hat{a}_{i, T} = a_{i, T}$ & Naive 1 & Naive 2 & Naive 3 \\
$\hat{a}_{i, T} = 0.1 \sum_{s=T-9}^T a_{i, s}$ & Naive 4 & Naive 5 & Naive 6\\
$\hat{a}_{i, T} = 0$ & Naive 7 & Naive 8 & Naive 9 \\
\hline
\end{tabular}}
\end{center}
\caption{Classification of naive forecast models by future acceleration and steering angle estimators.}
\label{tableNaive}
\end{table}

@@

Eucliedean error for each model, averaged across test set vehicles at forecast horizons of one, two, and three seconds ($h = 10, 20$ and $30$) are provided in Figure \ref{fig:PredError}. Naive model performance is split into three groups according to the choice of acceleration forecast, and for a given $\hat{a}_{i, T}$ the choice of $\hat{\delta}_{i, T}$ has not substantially changed the forecast error. The best performing Naive model at each forecast horizon is Naive 8, however at all forecast horizons the error for each implementation of each of the time series models are significantly improved relative to every naive model. The average forecast Euclidean error for Naive 8 and each of the time series models at each of these three forecast horizons is presented in Table \ref{table:meanError}.
\\

Figure \ref{fig:PredErrorZ} truncates Figure \ref{fig:PredError} to include only the time series models, each of which produces a forecast with a similar amount of error that does not increase substantially as the forecast horizon increases. This implies that there is no systematic increase or decrease in MAP forecast accuracy for choosing either the homogenous or introducing heterogeneity with the IH and CH models, however the UVB inference implementation of the IH model has increased the forecast error relative to MCMC and standard VB. The UVB implementation of the CH model produces forecast errors that are competitive with MCMC and standard VB, and so inference discussion will focus on the CH model rather than the IH model. 
\\

@@

The CH model allows each vehicle in the test to have a posterior variance distribution that is unique to that vehicle, while the homogenous model must share the same posterior variance distribution across different vehicles. Differences in the distribution of either variance parameter has little impact on the location of the maximum of $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$, but can drastically affect the precision of this forecast density, causing systematic differences in the level of uncertainty associated with the forecasts that can be measured by the logscore.
\\

Figure \ref{fig:posVarMean} displays plots of density of posterior means for each of acceleration and steering angle resulting from RWMH-MCMC sampling with $T = 450$. The homogenous model posterior means are indicated with vertical red lines. As in Figure \ref{fig:HierSingleKDE}, there is a large amount of variation in the distribution of variances for the test set, many of which are lower than the homogenous model: the homogenous posterior means are greater than $80\%$ and $91.5\%$ of CH posterior means for acceleration and the steering angle respectivley. 
\\

Figure \ref{fig:varSplit} shows a detailed breakdown of the median increase in predictive logscore obtained from using the CH model with UVB inference over the homogenous model at the three second forecast horizon. Vehicles are split into bivariate quintiles according to their posterior variance means for acceleration along the x-axis and angle along the y-axis. The upper left panels correspond to vehicles in the lower posterior variance quintiles, showing clear benefits for CH model trajectory forecasts relative to the homogenous model. However the median difference is zero, or even negative, for forecasts of drivers in the fifth quintiles. This paper argues that heterogenous modelling is still useful in this scenario, as the individual vehicle variance distributions allow identification of vehicles with high variance, which could display potentially more erratic and dangerous behaviour.
\\

@@

The benefits of heterogenous models is only useful if the posterior distribution $p(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$, or its variational approximation $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$ can be inferred in the short time frames demanded by self-driving vehicles. Figure \ref{fig:timing} demonstrates the average computation time required for standard VB (Black) and UVB (coloured), where UVB is first fit to $100$ observations then repeatedly updated every $T-S$ observations for different values of $T-S$. The compuation time for standard VB increases linearly with $T$, while UVB is constant as $T$ increases; and hence posterior inference utilising UVB to condition posterior distributions on additional data is feasible. Parallelising the $M$ Monte-Carlo samples per iteration can further reduce computation time, allowing for shorter update periods or more complex models. 
\\

As UVB involves replacing the posterior distribution at update with its variational approximation, the error introduced by UVB depends on how closely the variational approximation can represent the true posterior distribution. The left panel for Figure \ref{fig:updateCost} displays boxplots of the difference in logscore between UVB and standard VB at the three second forecast horizon for the CH model, and demonstrates that UVB in this case does not decrease the forecast logscores, as the difference between the logscores obtained by UVB and standard VB are distributed close to zero for several ranges of $T$. The right panel displays analogous boxplots for the IH model, where the UVB logscores tend to be lower than the standard VB logscores, matching the increased Euclidean error found for the IH model with UVB inference in Figure \ref{fig:PredErrorZ}. An explanation for this is given by the form of the approximating distribution used for the IH model, the multivariate normal distribution may not adequately approximate the posterior distribution, while the extra flexibilitiy of the mixture multivariate normal distribution for the CH model was sufficient to replace the posterior.

@@

\section{Discussion}
\label{sec:disc}
This paper demonstrates that the use of auto-regressive models time series models drastically improves trajectory forecasts when compared to the competing models.  These models parameterise the dynamics of a driver's actions, with a range of parameter posterior means providing evidence of heterogeneity between the drivers of different vehicles. Incorpororating this heterogeneity through a hierarchical model has improved the forecast logscores, and hence reduced uncertainty, but requires an approximate inference scheme implemented with UVB. 
\\
